Variational Inference for Inverse Reinforcement Learning with Gaussian Processes

### 2019-03-30 mpaper/paper.tex
Covariance experiment with cliques!
Describing the new plots and a few additions to the presentation

\section{Evaluation} \label{sec:evaluation}
% showing deep understanding

In order to fully understand the model's behaviour, we focus on a three-state
MDP where the agent can deterministically move from any state to any other
state. More formally, we set $\mathcal{S} = \{ s_1, s_2, s_3 \}$, $\mathcal{A} =
\{ a_1, a_2 \}$,
\begin{align*}
  &\mathcal{T}(s_1, a_1, s_2) = 1, \quad \mathcal{T}(s_1, a_2, s_3) = 1, \\
  &\mathcal{T}(s_2, a_1, s_1) = 1, \quad \mathcal{T}(s_1, a_2, s_3) = 1, \\
  &\mathcal{T}(s_3, a_1, s_1) = 1, \quad \mathcal{T}(s_1, a_2, s_2) = 1,
\end{align*}
all other values of $\mathcal{T}$ to zero, and $\gamma = 0.9$. We
also set the inducing points to be equal to the three states in $\mathcal{S}$,
add a single feature $f : \mathcal{S} \to \mathbb{R}$ such that
\[
  f(s_1) = 1, \quad f(s_2) = 2, \quad f(s_3) = 3,
\]
and create two demonstrations $\zeta_1 = \{ (s_1, a_1) \}$ and $\zeta_2 = \{
(s_3, a_2) \}$ that correspond to moving from $s_1$ and $s_3$ to $s_2$.
Therefore, we would expect the reward of $s_2$ to be higher than the other two
rewards in order to reflect this.

@@

We plot how $\mathcal{L}$ as well as policies $\pi(a_1 \mid s_1)$, $\pi(a_2 \mid
s_3)$, and $\pi(a_1 \mid s_2)$ converge over a number of iterations in Figure
\ref{fig:convergence1}. The first two policies correspond to actions taken in
the set of demonstrations $\mathcal{D}$, so we would expect to see their
probabilities converge to values near $1$. The third policy, however, has no
data in $\mathcal{D}$ to guide its value, so the maximum causal entropy
framework would put the probability at around $0.5$. The first thing to note is
that while $\mathcal{L}$ converges in just a few iterations, policies continue
to improve for much longer. In addition, all policies behave as expected:
$\pitwo$ stays around $0.5$, while both $\pione$ and $\pithree$ keep increasing.
Although they stay at around $0.75$ (and not 1), this may be optimal for the
maximum causal entropy model (and this question will be addressed in more detail
with a different experiment).

@@

\begin{figure}
  \centering
  \begin{tikzpicture}
    \csvreader[
    no head,
    column count=11
    ]{data/covariance_and_policy_transformed.csv}{1=\x, 2=\y, 3=\Soneone, 4=\Stwotwo, 5=\Sthreethree, 6=\Stwoone, 7=\Sthreeone, 8=\Sthreetwo, 9=\ponetwo, 10=\ptwoone, 11=\pthreetwo}{%
      \begin{axis}[
        ybar=0,
        symbolic x coords={a,b,c},
        ticks=none,
        ymax=1,
        ymin=0,
        x=20,
        bar width=20,
        enlarge x limits=0.25,
        at={(350*\x, 110*\y)},
        height=100,
        bar shift=0pt,
        legend style={
          legend columns=-1,
          at={(1.7,-0.4)}
        }
        ]
        \addplot coordinates {(a, \ponetwo)};
        \addplot coordinates {(b, \ptwoone)};
        \addplot coordinates {(c, \pthreetwo)};
        \draw[dashed] (-50,50) -- (270,50);
        \ifthenelse{\x=2 \AND \y=1}{
          \legend{$\pione$,$\pitwo$,$\pithree$};
        }{}
      \end{axis}
    }
    \draw [<->] (2.3, 8.5) node (yaxis) [above] {$\zeta_2$}
    |- (10, 2) node (xaxis) [right] {$\zeta_1$};
    \foreach \x in {1, 2, 3} {
      \node (x\x) at (1+2.5*\x, 1.7) {$\x$};
    }
    \foreach \y in {1, 2, 3} {
      \node (y\y) at (2.1, 1+2.1*\y) {$\y$};
    }
  \end{tikzpicture}
  \caption{Changes in policies after adding more demonstrations. The $x$-axis
    ($y$-axis) shows how many copies of $\zeta_1$ ($\zeta_2$) are in
    $\mathcal{D}$. Each plot shows the values of three policies after 300
    iterations (averaged out over 10 runs). The bottom (top) of each bar plot
    corresponds to probability 0 (1), and dashed lines mark probability
    0.5.}
  \label{fig:policies}
\end{figure}

@@

\begin{figure}[!ht]
  \centering
  \begin{tikzpicture}[
    /pgfplots/colormap/violet,
    ellipse2/.style={
      draw,
      ellipse,
      /utils/exec={
        \pgfplotscolormapdefinemappedcolor{#1}},
      fill=mapped color},
    edge2/.style={
      ultra thick,
      /utils/exec={
        \pgfplotscolormapdefinemappedcolor{#1}},
      draw=mapped color}]
    \csvreader[
    no head,
    column count=11
    ]{data/median_covariance_transformed.csv}{1=\x, 2=\y, 3=\Soneone, 4=\Stwotwo, 5=\Sthreethree, 6=\Stwoone, 7=\Sthreeone, 8=\Sthreetwo}{%
      \node[ellipse2=\Soneone] (s1) at (2*\x, 2*\y) {$s_1$};
      \node[ellipse2=\Stwotwo] (s2) at (2*\x+0.5, 2*\y+1) {$s_2$};
      \node[ellipse2=\Sthreethree] (s3) at (2*\x+1, 2*\y) {$s_3$};
      \draw[edge2=\Stwoone] (s1) -- (s2);
      \draw[edge2=\Sthreeone] (s1) to (s3);
      \draw[edge2=\Sthreetwo] (s2) to (s3);
    }
    \draw [<->] (1.5, 7.5) node (yaxis) [above] {$\zeta_2$}
    |- (8, 1.5) node (xaxis) [right] {$\zeta_1$};
    \foreach \x in {1, 2, 3} {
      \node (x\x) at (0.5+2*\x, 1.3) {$\x$};
    }
    \foreach \y in {1, 2, 3} {
      \node (y\y) at (1.3, 0.5+2*\y) {$\y$};
    }
    \begin{axis}[
      colormap/violet,
      scale only axis,
      width=0pt,
      height=0pt,
      hide axis,
      colorbar horizontal,
      point meta min=-1,
      point meta max=1,
      xshift=40,
      yshift=30,
      colorbar style={
        width=6.5cm,
        xtick={-1,0,1},
        xlabel=normalised covariance,
      }]
      \addplot [draw=none] coordinates {(0,0)};
    \end{axis}
  \end{tikzpicture}
  \caption{Changes in reward covariances after adding more demonstrations. The
    meaning of external axes is the same as in Figure \ref{fig:policies}.
    Colours denote reward covariance values, e.g., the colour of node $s_1$
    denotes the variance of $r(s_1)$, and the colour of the edge between nodes
    $s_1$ and $s_2$ denotes the covariance between $r(s_1)$ and $r(s_2)$. The
    colours represent median covariances across 10 runs, normalised to the
    interval $[-1, 1]$ while preserving their positivity/negativity.}
  \label{fig:covariance}
\end{figure}

@@

Unfortunately, the resulting plot in Figure \ref{fig:covariance} shows no clear
pattern. The only worthwhile observation is that there are only two instances of
negative covariance, and both of them are between states $s_2$ and $s_3$.
Perhaps there is a good (currently unknown) reason for why adding more data does
not reduce variance, or perhaps this plot could benefit from a higher number of
iterations and more repeated experiments to ensure the accuracy of the results.
% or a different MDP, like a star

% EXTRA INFO:
% Discarding iterations where either $\bm\Sigma$ or $\bm\Gamma$ becomes near-singular.
%Diagonal and non-diagonal entries of the covariance matrix are normalised to
%$[0, 1]$ separately, so as to emphasize the variation in each. There were no
%negative entries of the covariance matrix after convergence.
%More data creates bigger gradients---need smaller step size. Dynamic step sizes
%don't work very well because they become zero after a big spike (and spikes do
%happen).
\section{Related Work} % 0.5-1 page

@@

\subsection{Further Work}

>>>>

\section{Experiments} \label{sec:evaluation}
% showing deep understanding

In order to fully understand the model's behaviour, we start with a three-state
MDP where the agent can deterministically move from any state to any other
state. More formally, we set $\mathcal{S} = \{ s_1, s_2, s_3 \}$, $\mathcal{A} =
\{ a_1, a_2 \}$,
\begin{align*}
  &\mathcal{T}(s_1, a_1, s_2) = 1, \quad \mathcal{T}(s_1, a_2, s_3) = 1, \\
  &\mathcal{T}(s_2, a_1, s_1) = 1, \quad \mathcal{T}(s_1, a_2, s_3) = 1, \\
  &\mathcal{T}(s_3, a_1, s_1) = 1, \quad \mathcal{T}(s_1, a_2, s_2) = 1,
\end{align*}
all other values of $\mathcal{T}$ to zero, and $\gamma = 0.9$. We
also set the inducing points to be equal to the three states in $\mathcal{S}$,
add a single feature $f : \mathcal{S} \to \mathbb{R}$ such that
\[
  f(s_1) = 1, \quad f(s_2) = 2, \quad f(s_3) = 3,
\]
and create two demonstrations $\zeta_1 = \{ (s_1, a_1) \}$ and $\zeta_2 = \{
(s_3, a_2) \}$ that correspond to moving from $s_1$ and $s_3$ to $s_2$.
Therefore, we would expect the reward of $s_2$ to be higher than the other two
rewards in order to reflect this.

\paragraph{Convergence}

@@

We plot how $\mathcal{L}$ as well as policies $\pi(a_1 \mid s_1)$, $\pi(a_2 \mid
s_3)$, and $\pi(a_1 \mid s_2)$ converge over a number of iterations in Figure
\ref{fig:convergence1}\footnote{In all experiments, we discard runs that
  result in a near-singular $\bm\Sigma$ or $\bm\Gamma$, repeatedly testing each
  combination of parameters as many times as necessary.}. The first two policies
correspond to actions taken in the set of demonstrations $\mathcal{D}$, so we
would expect to see their probabilities converge to values near $1$. The third
policy, however, has no data in $\mathcal{D}$ to guide its value, so the maximum
causal entropy framework would put the probability at around $0.5$. The first
thing to note is that while $\mathcal{L}$ converges in just a few iterations,
policies continue to improve for much longer. In addition, all policies behave
as expected: $\pitwo$ stays around $0.5$, while both $\pione$ and $\pithree$
keep increasing. Although they stay at around $0.75$ (and not 1), this may be
optimal for the maximum causal entropy model (and this question will be
addressed in more detail with a different experiment).

@@

\paragraph{Adding More Data}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \csvreader[
    no head,
    column count=11
    ]{data/covariance_and_policy_transformed.csv}{1=\x, 2=\y, 3=\Soneone, 4=\Stwotwo, 5=\Sthreethree, 6=\Stwoone, 7=\Sthreeone, 8=\Sthreetwo, 9=\ponetwo, 10=\ptwoone, 11=\pthreetwo}{%
      \begin{axis}[
        ybar=0,
        symbolic x coords={a,b,c},
        ticks=none,
        ymax=1,
        ymin=0,
        x=20,
        bar width=20,
        enlarge x limits=0.25,
        at={(350*\x, 110*\y)},
        height=100,
        bar shift=0pt,
        legend style={
          legend columns=-1,
          at={(1.7,-0.4)}
        }
        ]
        \addplot coordinates {(a, \ponetwo)};
        \addplot coordinates {(b, \ptwoone)};
        \addplot coordinates {(c, \pthreetwo)};
        \draw[dashed] (-50,50) -- (270,50);
        \ifthenelse{\x=2 \AND \y=1}{
          \legend{$\pione$,$\pitwo$,$\pithree$};
        }{}
      \end{axis}
    }
    \draw [<->] (2.3, 8.5) node (yaxis) [above] {$\zeta_2$}
    |- (10, 2) node (xaxis) [right] {$\zeta_1$};
    \foreach \x in {1, 2, 3} {
      \node (x\x) at (1+2.5*\x, 1.7) {$\x$};
    }
    \foreach \y in {1, 2, 3} {
      \node (y\y) at (2.1, 1+2.1*\y) {$\y$};
    }
  \end{tikzpicture}
  \caption{Changes in policies after adding more demonstrations. The $x$-axis
    ($y$-axis) shows how many copies of $\zeta_1$ ($\zeta_2$) are in
    $\mathcal{D}$. Each plot shows the values of three policies after 300
    iterations (averaged out over ten runs). The bottom (top) of each bar plot
    corresponds to probability 0 (1), and dashed lines mark probability
    0.5.}
  \label{fig:policies}
\end{figure}

@@

\begin{figure}[!ht]
  \centering
  \begin{tikzpicture}[
    /pgfplots/colormap/violet,
    ellipse2/.style={
      draw,
      circle,
      /utils/exec={
        \pgfplotscolormapdefinemappedcolor{#1}},
      fill=mapped color},
    edge2/.style={
      ultra thick,
      /utils/exec={
        \pgfplotscolormapdefinemappedcolor{#1}},
      draw=mapped color}]
    \csvreader[
    no head,
    column count=11
    ]{data/median_covariance_transformed.csv}{1=\x, 2=\y, 3=\Soneone, 4=\Stwotwo, 5=\Sthreethree, 6=\Stwoone, 7=\Sthreeone, 8=\Sthreetwo}{%
      \node[ellipse2=\Soneone] (s1) at (2*\x, 2*\y) {$s_1$};
      \node[ellipse2=\Stwotwo] (s2) at (2*\x+0.5, 2*\y+1) {$s_2$};
      \node[ellipse2=\Sthreethree] (s3) at (2*\x+1, 2*\y) {$s_3$};
      \draw[edge2=\Stwoone] (s1) -- (s2);
      \draw[edge2=\Sthreeone] (s1) to (s3);
      \draw[edge2=\Sthreetwo] (s2) to (s3);
    }
    \draw [<->] (1.5, 7.5) node (yaxis) [above] {$\zeta_2$}
    |- (8, 1.5) node (xaxis) [right] {$\zeta_1$};
    \foreach \x in {1, 2, 3} {
      \node (x\x) at (0.5+2*\x, 1.3) {$\x$};
    }
    \foreach \y in {1, 2, 3} {
      \node (y\y) at (1.3, 0.5+2*\y) {$\y$};
    }
    \begin{axis}[
      colormap/violet,
      scale only axis,
      width=0pt,
      height=0pt,
      hide axis,
      colorbar horizontal,
      point meta min=-1,
      point meta max=1,
      xshift=40,
      yshift=30,
      colorbar style={
        width=6.5cm,
        xtick={-1,0,1},
        xlabel=normalised covariance,
      }]
      \addplot [draw=none] coordinates {(0,0)};
    \end{axis}
  \end{tikzpicture}
  \caption{Changes in reward covariances after adding more demonstrations. The
    meaning of external axes is the same as in Figure \ref{fig:policies}.
    Colours denote reward covariance values, e.g., the colour of node $s_1$
    denotes the variance of $r(s_1)$, and the colour of the edge between nodes
    $s_1$ and $s_2$ denotes the covariance between $r(s_1)$ and $r(s_2)$. The
    colours represent median covariances across ten runs, normalised to the
    interval $[-1, 1]$ while preserving their positivity/negativity.}
  \label{fig:covariance}
\end{figure}

@@

Unfortunately, the resulting plot in Figure \ref{fig:covariance} shows no clear
pattern, suggesting that adding more of the same demonstrations does not result
in lower covariances. The only other observation from this plot is that there
are only two instances of negative covariance, and both of them are between
states $s_2$ and $s_3$.

\paragraph{Structure versus Randomness}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \foreach \i in {0,...,9}
    \node[draw,circle] (v\i) at (360/10 * \i:3cm) {$s_{\i}$};
    \foreach \i in {0,...,9}
    \foreach \j in {0,...,\i}
    \draw[gray] (v\i) -- (v\j);

    \draw[->,thick,blue] (v9) -- (v2);
    \draw[->,thick,blue] (v7) -- (v8);
    \draw[->,thick,blue] (v9) -- (v4);
    \draw[->,thick,blue] (v3) -- (v0);
    \draw[->,thick,blue] (v2) -- (v4);
    \draw[->,thick,blue] (v4) -- (v6);
    \draw[->,thick,blue] (v6) -- (v5);

    \begin{scope}[yshift=-7cm]
    \foreach \i in {0,...,9}
    \node[draw,circle] (v\i) at (360/10 * \i:3cm) {$s_{\i}$};
    \foreach \i in {0,...,9}
    \foreach \j in {0,...,\i}
    \draw[gray] (v\i) -- (v\j);

    \draw[->,thick,blue] (v3) -- (v0);
    \draw[->,thick,blue] (v4) -- (v0);
    \draw[->,thick,blue] (v6) -- (v0);
    \draw[->,thick,blue] (v2) -- (v0);
    \draw[->,thick,blue] (v5) -- (v0);
    \draw[->,thick,blue] (v8) -- (v0);
    \draw[->,thick,blue] (v9) -- (v0);
    \end{scope}
  \end{tikzpicture}
  \caption{The clique MDP with two sets of demonstrations denoted by blue arrows.}
  \label{fig:cliques}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{figures/boxplots}
  \caption{Box plots of the difference between (absolute values of) reward
    covariances in random and semi-structured scenarios from Figure
    \ref{fig:cliques}, averaged out over 100 runs and grouped into four key
    categories. `First state' refers to the variance of $r(s_0)$, the reward of
    the state targeted by all demonstrations in the semi-structured case. `Other
    states' are variances of other states of the MDP, i.e., diagonal entries of
    the covariance matrix. `Incident edges' refers to covariances between
    $r(s_0)$ and all other states, i.e., the first row or column of the
    covariance matrix. Finally, `other edges' is the category for the remaining
    covariances between pairs of rewards.}
  \label{fig:covariance_boxplots}
\end{figure}

If having more copies of the same demonstrations does not affect covariances, we
can test if covariances tend to be lower in situations that exhibit more
structure, i.e., demonstrations provide a clearer picture of which states are
more valuable. For this, we consider a completely-deterministic MDP with ten
states and nine actions that allow the agent to move from any state to any other
state---the \emph{clique MDP}. We denote the states by $\mathcal{S} = \{ s_i
\}_{i=0}^9$, and, similarly to the previous example, set up a single feature $f$
such that $f(s_i) = i$ for $i = 0, \dots, 9$. Finally, $\gamma = 0.9$ as before.
We consider two scenarios, with 100 randomly generated demonstrations in each.
The scenarios are visualised in Figure \ref{fig:cliques}, although we only show
a small subset of demonstrations. In the first scenario, we draw both the
starting state and the action uniformly at random. In the second scenario, we
draw the starting state uniformly from $\mathcal{S} \setminus \{ s_0 \}$, and
the action always points to $s_0$. We would expect at least variances of state
rewards to be lower in the second scenario in order to reflect the more
structured, certain behavioural pattern expressed by the demonstrations.

Indeed, Figure \ref{fig:covariance_boxplots} shows exactly that. The first two
box plots show that in roughly $3/4$ of the runs reward variances were higher in
the random scenario\footnote{Note that outliers were removed from the plot.
  However, all outliers were positive.}. Interestingly, the same applies to
covariances.

% TODO: conclude the experiments

\section{Related Work} % 0.5-1 page

@@

\subsection{Further Work}
%TODO: how optimal policies depend on the structure of the mdp and available
%data (i.e., why don't our policies converge to 1)
